---
title: "Introduction to gsbundle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gsbundle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE   # network calls; set to TRUE locally if desired
)
```

gsbundle provides three core capabilities for gene-centric analysis in R:

1. **Gene set retrieval** — download and cache gene sets from KEGG,
   Reactome, Gene Ontology, MSigDB, and CORUM
2. **Gene ID conversion** — translate between Entrez, Symbol, Ensembl,
   and UniProt identifiers, within and across organisms
3. **Enrichment analysis** — run hypergeometric, over-representation, or
   GSEA-style tests with optional DBSCAN-based redundancy filtering

All data are downloaded from public databases on first use and stored in
a local cache directory, so subsequent calls work offline and are fast.
The cache location is fully configurable — see
[Section 6: Configuring the cache directory](#configuring-the-cache-directory).

```{r}
library(gsbundle)
```

---

## 1. Retrieving gene sets

`gsGetter()` is the main entry point for fetching gene–pathway mappings.
It returns a three-column data frame (`Gene`, `PathwayID`, `PathwayName`)
ready to pass directly to `runEnrich()`.

```{r}
# KEGG and Reactome pathways (human)
gene2path <- gsGetter(type = "KEGG+REACTOME", organism = "hsa")
head(gene2path)
```

### Available databases and shorthands

| `type` value | Databases included |
|---|---|
| `All` | Everything below |
| `Pathway` | KEGG, REACTOME, C2_CP_PID, C2_CP_BIOCARTA, C2_CP_WIKIPATHWAYS, C2_CP_KEGG_MEDICUS |
| `GO` | GOBP, GOCC, GOMF |
| `Complex` | CORUM protein complexes |
| `MSIGDB` | All MSigDB collections (C1–C8, H) |
| Individual | `KEGG`, `REACTOME`, `GOBP`, `GOCC`, `GOMF`, `CORUM`, `C1`…`C8`, `H` |

Combine any types with `+`:

```{r}
# Biological Process GO terms and KEGG only
go_kegg <- gsGetter(type = "GOBP+KEGG", organism = "hsa")

# Filter to gene sets with 15–500 members
kegg_filtered <- gsGetter(type = "KEGG", organism = "hsa", limit = c(15, 500))

# Mouse gene sets
gene2path_mmu <- gsGetter(type = "KEGG+REACTOME", organism = "mmu")
```

MSigDB requires a local `.db` SQLite file (available from
[msigdb.org](https://www.gsea-msigdb.org/gsea/msigdb)):

```{r}
msigdb_gs <- gsGetter(type = "H+C2", organism = "hsa",
                      msigdb.path = "/path/to/msigdb.v2024.1.Hs.db")
```

### Custom GMT files

`gsGetter()` accepts a path to any GMT-format file via `gmtpath`. The
built-in `ReadGMT()` helper can also be used directly if you need to
parse a GMT file without the rest of `gsGetter()`'s logic.

```{r}
# Via gsGetter (respects limit=)
custom <- gsGetter(gmtpath = "my_gene_sets.gmt", limit = c(10, 500))

# Via ReadGMT directly
gene2path_custom <- ReadGMT("my_gene_sets.gmt", limit = c(10, 500))
```

### Updating cached data

Gene set databases are downloaded once. To refresh from source:

```{r}
# Force re-download of KEGG and Reactome
gsGetter(type = "KEGG+REACTOME", organism = "hsa", update = TRUE)

# Or call retrieve_gs() directly to pre-populate the cache
retrieve_gs(type = c("KEGG", "REACTOME", "GO"), organism = "hsa")
```

### Checking cached database versions

`gs_versions()` reads the metadata file recorded after each download and
returns a data frame showing which version of each database was cached and
when it was downloaded.

```{r}
gs_versions()
#>   database organism       version  downloaded        source_url
#> 1     kegg      hsa  Release 113  2025-01-15  https://rest.kegg.jp/...
#> 2  ensembl      hsa          114  2025-01-15  https://ftp.ensembl.org/...
#> 3       go      hsa  (fetched..) 2025-01-15  https://ftp.ncbi.nlm.nih.gov/...
```

---

## 2. Gene ID conversion

`TransGeneID()` converts a character vector of gene identifiers between
formats. Built-in mappings (no internet required after first use) cover
`Symbol`, `Entrez`, `Ensembl`, and `UniProt`. Additional formats are
available via `biomaRt`.

The function returns a **named character vector** — names are the original
input identifiers, values are the converted IDs. This makes it easy to
propagate identifiers through a pipeline without losing track of the
input→output mapping.

```{r}
# Symbol → Entrez (default)
TransGeneID(c("TP53", "BRCA1", "EGFR"), organism = "hsa")
#> TP53  BRCA1   EGFR
#> "7157" "672" "1956"

# Symbol → UniProt
TransGeneID(c("TP53", "BRCA1"), toType = "uniprot", organism = "hsa")

# Entrez → Ensembl
TransGeneID(c("7157", "672"), fromType = "Entrez", toType = "Ensembl",
            organism = "hsa")
```

### One-to-many mappings

By default, `unique = TRUE` returns only the first match per input gene.
Set `unique = FALSE` to get a data frame of all one-to-many mappings:

```{r}
# Returns a data frame with all Ensembl IDs per Entrez ID
TransGeneID(c("7157", "672"), fromType = "Entrez", toType = "Ensembl",
            organism = "hsa", unique = FALSE)
```

### Cross-organism mapping

```{r}
# Mouse symbol → human symbol
TransGeneID("Trp53", toType = "Symbol", fromOrg = "mmu", toOrg = "hsa")

# Mouse Entrez → human Entrez
TransGeneID("22059", fromType = "Entrez", toType = "Entrez",
            fromOrg = "mmu", toOrg = "hsa")
```

Cross-organism mapping for `Symbol` and `Entrez` uses built-in annotations
from MGI and NCBI HomoloGene (downloaded via `getOrtAnn()`). For other ID
types, `biomaRt` is used automatically.

### Updating gene annotations

The underlying annotation files are cached locally. Pass `update = TRUE` to
re-download from NCBI and Ensembl:

```{r}
ann <- getGeneAnn("hsa", update = TRUE)
head(ann$Gene)     # entrez, symbol, ensembl, synonyms, ...
head(ann$Protein)  # uniprot, symbol, synonyms
```

---

## 3. Enrichment analysis

`runEnrich()` wraps `clusterProfiler` and works directly with the output
of `gsGetter()`. The input `geneList` is a **named numeric vector** —
names are gene IDs (Entrez by default) and values are scores such as
log~2~ fold-changes, *t*-statistics, or –log~10~ *p*-values.

```{r}
if (!requireNamespace("clusterProfiler", quietly = TRUE))
    BiocManager::install("clusterProfiler")

gene2path <- gsGetter(type = "KEGG+GOBP", organism = "hsa")

# Example: a named vector of scores (replace with real data)
set.seed(42)
geneList <- setNames(rnorm(300), sample(gene2path$Gene, 300))
```

### Hypergeometric test (over-representation)

The hypergeometric test (`"HGT"`) asks whether a set of genes of interest
is over-represented in each pathway relative to a background universe.
The gene *names* form the hit list; the numeric values contribute to a
normalised enrichment score (`NES`) and `strength` (log~10~ enrichment
ratio) that are appended to the result.

```{r}
res_hgt <- runEnrich(geneList, gene2path,
                     method       = "HGT",
                     pvalueCutoff = 0.05)
head(as.data.frame(res_hgt))
```

The result data frame contains: `ID`, `Description`, `NES`, `meanVal`,
`strength`, `pvalue`, `p.adjust`, `GeneRatio`, `BgRatio`, `geneID`
(Entrez IDs), `geneName` (symbols), and `Count`.

### Over-representation test

`"ORT"` is equivalent to `clusterProfiler::enricher()` with no NES
annotation:

```{r}
res_ort <- runEnrich(geneList, gene2path,
                     method       = "ORT",
                     pvalueCutoff = 0.05)
```

### GSEA

GSEA ranks all genes by their score and tests whether pathway members
cluster at the top or bottom of the ranked list.

```{r}
res_gsea <- runEnrich(geneList, gene2path,
                      method       = "GSEA",
                      pvalueCutoff = 0.25)
```

### Input in gene symbols

If your `geneList` is named by gene symbols rather than Entrez IDs,
specify `keytype`:

```{r}
symbol_list <- setNames(rnorm(100),
                        sample(c("TP53", "MYC", "EGFR", "BRCA1"), 100,
                               replace = TRUE))
res <- runEnrich(symbol_list, gene2path, keytype = "Symbol", method = "HGT")
```

### Custom background universe

By default the background is all genes in `gene2path` combined with the
input genes. Supply your own universe with the `universe` argument:

```{r}
bg <- unique(gene2path$Gene)   # all genes in the gene set collection
res <- runEnrich(geneList, gene2path, method = "HGT",
                 pvalueCutoff = 0.05, universe = bg)
```

---

## 4. Reducing redundant pathways

Enrichment results from large databases often contain many overlapping
pathways. gsbundle offers two complementary strategies.

### DBSCAN clustering

DBSCAN groups pathways by Jaccard similarity of their gene sets. Requires
the `dbscan` package. The key tuning parameter is `eps`: larger values
merge more pathways into the same cluster (coarser grouping); smaller
values produce finer grouping.

```{r}
# Add a cluster column without removing any rows
res_clust <- runEnrich(geneList, gene2path, method = "HGT",
                       pvalueCutoff = 0.05,
                       cluster = TRUE, eps = 1.2, minpts = 2)

# Pick the top pathway per cluster yourself:
library(dplyr)
top_per_cluster <- as.data.frame(res_clust) |>
    slice_min(pvalue, n = 1, by = cluster)
```

```{r}
# Or let runEnrich keep only the best representative per cluster
res_filt <- runEnrich(geneList, gene2path, method = "HGT",
                      pvalueCutoff = 0.05,
                      filter = TRUE, eps = 1.2, minpts = 2)
```

`cluster.dbscan()` can also be called directly on an existing result data
frame if you want to add clustering after the fact:

```{r}
clusters <- cluster.dbscan(as.data.frame(res_hgt), eps = 1.2, minpts = 2)
```

### Jaccard cutoff filter

`EnrichedFilter()` applies a simpler pairwise Jaccard cutoff and removes
the lower-ranked member of any pair that exceeds it:

```{r}
filtered_df <- EnrichedFilter(res_hgt, cutoff = 0.8)
```

---

## 5. Utility functions

### Read a GMT file

```{r}
gene2path_custom <- ReadGMT("my_gene_sets.gmt", limit = c(10, 500))
```

### Format MSigDB pathway names

MSigDB gene set names are stored in `SCREAMING_SNAKE_CASE`. `format_gs_name()`
converts them to readable sentence-case labels, handling Greek letters and
common acronyms:

```{r}
format_gs_name(c("TNFA_SIGNALING_VIA_NFKB",
                 "PI3K_AKT_MTOR_SIGNALING",
                 "KRAS_SIGNALING_DN",
                 "PANCREAS_BETA_CELLS"))
#> [1] "TNF-α signaling via NF-κB"
#> [2] "PI3K-AKT-mTOR signaling"
#> [3] "KRAS signaling down"
#> [4] "Pancreas β cells"
```

This is applied automatically inside `gsGetter()` for MSigDB collections.

### Organism helpers

```{r}
getOrg("human")   # $org = "hsa",  $pkg = "org.Hs.eg.db"
getOrg("mmu")     # $org = "mmu",  $pkg = "org.Mm.eg.db"
```

### Retrieve raw gene annotations

```{r}
# Gene and protein annotations for human (downloads once, then cached)
ann <- getGeneAnn("hsa")
head(ann$Gene)     # columns: entrez, symbol, ensembl, synonyms, ...
head(ann$Protein)  # columns: uniprot, symbol, synonyms

# Ortholog table: mouse → human
orth <- getOrtAnn(fromOrg = "mmu", toOrg = "hsa")
head(orth)
```

---

## 6. Configuring the cache directory

All cached files (gene set RDS files, gene annotation files, and the
`versions.json` metadata record) are stored in a single directory. The
location is resolved in this priority order:

| Priority | Mechanism | Scope |
|---|---|---|
| 1 (highest) | `cache.dir` argument | Single function call |
| 2 | `options("gsbundle.cache")` | Entire R session |
| 3 (default) | `tools::R_user_dir("gsbundle", "cache")` | Per-user system default |

The system default is typically:

- **Linux/macOS**: `~/.cache/R/gsbundle/`
- **Windows**: `%LOCALAPPDATA%\R\Cache\R\gsbundle\`

### Per-call override

Pass `cache.dir` directly to `gsGetter()` or `retrieve_gs()` to redirect
a single call:

```{r}
# Read from (and download to) a specific directory for this call only
gene2path <- gsGetter(type = "KEGG+GOBP", organism = "hsa",
                      cache.dir = "/data/shared/gsbundle_cache")
```

### Session-wide option

Set `options("gsbundle.cache")` once at the top of a script and all
subsequent calls will use that directory:

```{r}
options(gsbundle.cache = "/data/shared/gsbundle_cache")

gene2path <- gsGetter(type = "KEGG+GOBP", organism = "hsa")
ann       <- getGeneAnn("hsa")
orth      <- getOrtAnn("mmu", "hsa")
gs_versions()
```

### Persistent default via `.Rprofile`

To apply the setting automatically every time R starts, add the option to
`~/.Rprofile` (user-wide) or `.Rprofile` in a project directory:

```r
# ~/.Rprofile  (or project .Rprofile)
options(gsbundle.cache = "/data/shared/gsbundle_cache")
```

### Shared / HPC cache

On a computing cluster or shared workstation it is efficient to download
data once into a shared location that all users or jobs can read:

```{r}
# --- Step 1: populate the cache (done once, e.g. by a data manager) ---
options(gsbundle.cache = "/data/shared/gsbundle_cache")

retrieve_gs(type = c("KEGG", "REACTOME", "GO", "CORUM"),
            organism = "hsa")
retrieve_gs(type = c("KEGG", "REACTOME", "GO"),
            organism = "mmu")
getGeneAnn("hsa")
getGeneAnn("mmu")

# --- Step 2: point every user/job at the shared cache ---
# (add to ~/.Rprofile or set at the top of each script)
options(gsbundle.cache = "/data/shared/gsbundle_cache")

gene2path <- gsGetter(type = "KEGG+GOBP", organism = "hsa")
```

This avoids redundant downloads and ensures all analyses use the same
database snapshot.

---

## Session info

```{r, eval = TRUE}
sessionInfo()
```
